---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# endogenr

<!-- badges: start -->
<!-- badges: end -->

The goal of endogenr is to make it easy to simulate dynamic systems from regression models, mathematical equations, and exogenous inputs (either based on a stochastic distribution, or given by some data). We assume a panel-data structure, with two variables denoting the time and the space dimensions.

The simulator works by identifying the dependency graph of the models added to the system, and deriving the order of calculation based on this graph.

It works with parallel simulation using the future::multisession approach.

## Installation

You can install the development version of endogenr from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("prio-data/endogenr")
```

## Example

```{r example}
library(endogenr)

train_start = 1970; test_start = 2010; timevar = "year"; groupvar = "gwcode"
df <- example_data
train <- df[time >= train_start & time < test_start, env = list(time = timevar)]

e1 <- gdppc_grwt ~ dplyr::lag(yjbest) + dplyr::lag(gdppc_grwt) + dplyr::lag(log(gdppc)) + dplyr::lag(psecprop) + dplyr::lag(zoo::rollmean(gdppc_grwt, k = 3, fill = NA, align = "right"))
c1 <- yjbest ~ dplyr::lag(yjbest) + dplyr::lag(log(gdppc)) + dplyr::lag(log(population)) + dplyr::lag(psecprop) + dplyr::lag(dem) + dplyr::lag(gdppc_grwt) + dplyr::lag(zoo::rollmean(yjbest, k = 5, fill = NA, align = "right"))
d1 <- dem ~ dplyr::lag(dem) + dplyr::lag(gdppc_grwt) + dplyr::lag(log(gdppc)) + dplyr::lag(yjbest) + dplyr::lag(psecprop) + dplyr::lag(zoo::rollmean(dem, k = 3, fill = NA, align = "right"))

m1 <- StochasticStaticModel$new(~gdppc_grwt, "t_ls", fit_args = list(start = list(df = 1, mu = mean(train$gdppc_grwt), sigma = sd(train$gdppc_grwt))), censor = list("upper" = 0.8, "lower" = -0.8))
m2 <- DeterministicModel$new(gdppc ~ I(abs(dplyr::lag(gdppc)*(1+gdppc_grwt))))
m3 <- ExogenModel$new(~population, exogendata = df)
m4 <- DeterministicModel$new(gdp ~ I(abs(gdppc * population)))
m5 <- StochasticStaticModel$new(~round(yjbest), "nbinom", fit_args = list(discrete = TRUE, method = "mme"))
m6 <- LinearRegressionModel$new(d1, boot = T)
m7 <- ExogenModel$new(~psecprop, exogendata = df)

models <- list(m1, m2, m3, m4, m5, m6, m7)

simulator <- EndogenousSystem$new(
  df, 
  timevar = timevar, 
  groupvar = groupvar,
  train_start = train_start,
  test_start = test_start, 
  min_window = 10
)

models <- list(m1, m2, m3, m4, m5, m6, m7)
for(model in models){
  simulator$add_model(model)
}

simulator$simulate(horizon = 12, nsim = 2, parallel = F)
simulator$plot("gdppc_grwt", units = c(2, 101, 365, 710, 475))
```
