---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# endogenr

<!-- badges: start -->
<!-- badges: end -->

The goal of endogenr is to make it easy to simulate dynamic systems from regression models, mathematical equations, and exogenous inputs (either based on a stochastic distribution, or given by some data). We assume a panel-data structure, with two variables denoting the time and the space dimensions.

The simulator works by identifying the dependency graph of the models added to the system, and deriving the order of calculation based on this graph.

It works with parallel simulation using the future::multisession approach.

## Installation

You can install the development version of endogenr from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("prio-data/endogenr")
# alternatively with
renv::install("prio-data/endogenr")
```

You can also clone (download) the repository, open it as a project in RStudio, find the "Build" tab, and press "Install".

## Example

```{r example}
library(endogenr)
library(fable)
library(dplyr)
df <- endogenr::example_data
df <- tsibble::as_tsibble(df, key = "gwcode", index = "year")

e1 <- gdppc_grwt ~ lag(yjbest) + lag(gdppc_grwt) + lag(log(gdppc)) + lag(psecprop) + lag(zoo::rollmean(gdppc_grwt, k = 3, fill = NA, align = "right"))
c1 <- yjbest ~ lag(yjbest) + lag(log(gdppc)) + lag(log(population)) + lag(psecprop) + lag(dem) + lag(gdppc_grwt) + lag(zoo::rollmean(yjbest, k = 5, fill = NA, align = "right"))
d1 <- dem ~ lag(dem) + lag(gdppc_grwt) + lag(log(gdppc)) + lag(yjbest) + lag(psecprop) + lag(zoo::rollmean(dem, k = 3, fill = NA, align = "right"))

model_system <- list(
  build_model("deterministic",formula = gdppc ~ I(abs(lag(gdppc)*(1+gdppc_grwt)))),
  build_model("deterministic", formula = gdp ~ I(abs(gdppc*population))),
  build_model("parametric_distribution", formula = ~gdppc_grwt, distribution = "norm"),
  build_model("linear", formula = c1, boot = "resid"),
  build_model("univariate_fable", formula = dem ~ error("A") + trend("N") + season("N"), method = "ets"),
  build_model("exogen", formula = ~psecprop),
  build_model("exogen", formula = ~population)
)

simulator_setup <- setup_simulator(models = model_system,
                                   data = df,
                                   train_start = 1970,
                                   test_start = 2010,
                                   horizon = 12,
                                   groupvar = "gwcode",
                                   timevar = "year",
                                   inner_sims = 2,
                                   min_window = 10)

set.seed(42)
res <- simulate_endogenr(nsim = 2, simulator_setup = simulator_setup, parallel = F)

# Example that you can back-transform variables you have simulated as transformed variables
scaled_logit <- function(x, lower=0, upper=1){
  log((x-lower)/(upper-x))
}
inv_scaled_logit <- function(x, lower=0, upper=1){
  (upper-lower)*exp(x)/(1+exp(x)) + lower
}
my_scaled_logit <- fabletools::new_transformation(scaled_logit, inv_scaled_logit)

yj <- scales::transform_yj(p = 0.4) # This was transformed using lambda 0.4 when creating the data.
# Back-transform
res <- res |> dplyr::mutate(v2x_libdem = inv_scaled_logit(dem),
                            best = yj$inverse(yjbest))

res <- tsibble::tsibble(res, key = c(simulator_setup$groupvar, ".sim"), index = simulator_setup$timevar) |>
  dplyr::filter(year >= simulator_setup$test_start)

acc <- get_accuracy(res, "gdppc_grwt", df)
acc |> summarize(across(crps:winkler, ~ mean(.x))) |> arrange(crps) |> knitr::kable()

plotsim(res, "gdppc", c(2, 20, 530), df)
plotsim(res, "gdppc_grwt", c(2, 20, 530), df)
plotsim(res, "v2x_libdem", c(2, 20, 530), df)
plotsim(res, "best", c(2, 20, 530), df)
```

## Spatial Lag

`endogenr` supports spatial lag variables in the simulation. The setup involves two steps: (1) computing the spatial lag for the historical data, and (2) registering a `spatial_lag` model in the model system so the lag is recomputed at each simulated time step.

### Step 1: Prepare the spatial weights and compute the historical lag

The convenience function `st_weights_from_sf` creates a neighborhood list and weights from an `sf` object. It defaults to queen contiguity. For more advanced weight schemes, use the `sfdep` package directly and supply the neighborhood list (`nb`), weights (`wt`), and the unit IDs in the order they appear in the spatial object.

Note that computing the spatial lag for the historical data can be fiddly when the panel is unbalanced or has missing observations. The example below filters the data to units present in the neighborhood structure before computing the lag.

```{r spatial-lag, eval=FALSE}
library(endogenr)
library(dplyr)

df <- endogenr::example_data
df <- tsibble::as_tsibble(df, key = "gwcode", index = "year")

# Load a map and filter to units present in the data
map <- poldat::cshp_gw_modifications(france_overseas = FALSE) |>
  dplyr::filter(end == as.Date("2019-12-31"))
map <- map |> dplyr::filter(gwcode %in% unique(df$gwcode))

# Build spatial weights (queen contiguity by default)
sf::sf_use_s2(FALSE)
neigh <- st_weights_from_sf(map, "gwcode", weights_args = list(allow_zero = TRUE))

# Compute the spatial lag of yjbest for each year in the historical data
df <- df |>
  dplyr::filter(gwcode %in% neigh$unit_ids) |>
  as_tibble() |>
  group_by(year) |>
  dplyr::mutate(
    sl_yjbest = sfdep::st_lag(yjbest, neigh$nb, neigh$wt, allow_zero = TRUE)
  )

df <- tsibble::tsibble(df, key = "gwcode", index = "year")
```

### Step 2: Add a `spatial_lag` model to the system

A `spatial_lag` model works like a deterministic model — it is a transformation applied at each simulated time step `t`. This ensures the spatial lag variable is updated as the underlying variable evolves during simulation.

One important consideration is **how the spatial lag variable enters other model formulas**. Using `yjbest ~ sl_yjbest` will not work because `sl_yjbest` is computed in the same period as `yjbest`, creating a circular dependency. Instead, use `lag(sl_yjbest)` so the conflict model uses the spatial lag from the previous period.

```{r spatial-lag-system, eval=FALSE}
e1 <- gdppc_grwt ~ lag(yjbest) + lag(gdppc_grwt) + lag(log(gdppc)) + lag(psecprop) +
  lag(zoo::rollmean(gdppc_grwt, k = 3, fill = NA, align = "right"))
c1 <- yjbest ~ lag(yjbest) + lag(sl_yjbest) + lag(log(gdppc)) + lag(log(population)) +
  lag(psecprop) + lag(dem) + lag(gdppc_grwt) +
  lag(zoo::rollmean(yjbest, k = 5, fill = NA, align = "right"))
d1 <- dem ~ lag(dem) + lag(gdppc_grwt) + lag(log(gdppc)) + lag(yjbest) + lag(psecprop) +
  lag(zoo::rollmean(dem, k = 3, fill = NA, align = "right"))

model_system <- list(
  # spatial_lag recomputes sl_yjbest from yjbest at each simulated t
  build_model("spatial_lag", formula = sl_yjbest ~ yjbest,
              nb = neigh$nb, wt = neigh$wt, unit_ids = neigh$unit_ids,
              island_default = 0),
  build_model("deterministic", formula = gdppc ~ I(abs(lag(gdppc) * (1 + gdppc_grwt)))),
  build_model("deterministic", formula = gdp ~ I(abs(gdppc * population))),
  build_model("parametric_distribution", formula = ~gdppc_grwt, distribution = "norm"),
  build_model("linear", formula = c1, boot = "resid"),
  build_model("univariate_fable", formula = dem ~ error("A") + trend("N") + season("N"),
              method = "ets"),
  build_model("exogen", formula = ~psecprop),
  build_model("exogen", formula = ~population)
)

simulator_setup <- setup_simulator(
  models = model_system,
  data = df,
  train_start = 1970,
  test_start = 2010,
  horizon = 12,
  groupvar = "gwcode",
  timevar = "year",
  inner_sims = 2,
  min_window = 10
)

set.seed(42)
res <- simulate_endogenr(nsim = 2, simulator_setup = simulator_setup, parallel = FALSE)
```
