---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# endogenr

<!-- badges: start -->
<!-- badges: end -->

The goal of `endogenr` is to make it easy to simulate dynamic systems from
regression models, mathematical equations, and exogenous inputs (either
based on a stochastic distribution, or given by some data). It assumes a
panel-data structure with two columns identifying the time and the unit
dimensions.

The simulator identifies the dependency graph of the models added to the
system and derives the order of calculation from that graph. Parallel
execution is opt-in via the `future` package.

## Installation

You can install the development version of `endogenr` from
[GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("prio-data/endogenr")
# alternatively with
renv::install("prio-data/endogenr")
```

You can also clone the repository, open it as a project in RStudio,
find the "Build" tab, and press "Install".

## Example

`setup_simulator()` accepts a plain `data.frame` or `data.table` — no
`tsibble` conversion is required. Formula RHS terms can use `lag()`,
`zoo::rollmean()`, or any function you define yourself (pass user functions
through the `globals` argument).

```{r example}
library(endogenr)
library(dplyr)
df <- endogenr::example_data

# Drop units with any NA in modelled outcomes over the training window 1970–2009.
required <- c("gdppc", "gdppc_grwt", "best", "v2x_polyarchy", "psecprop", "population")
train_window <- df[df$year >= 1970 & df$year <= 2009, ]
ok <- aggregate(train_window[, required],
                by = list(gwcode = train_window$gwcode),
                FUN = function(x) all(!is.na(x)))
keep <- ok$gwcode[apply(ok[, required], 1, all)]
df   <- df[df$gwcode %in% keep, ]
df$gdp <- df$gdppc * df$population  # derived outcome; pre-computed so the initial state is non-NA

c1 <- best ~ lag(best) + lag(log(gdppc)) + lag(log(population)) +
  lag(psecprop) + lag(v2x_polyarchy) + lag(gdppc_grwt) +
  lag(zoo::rollmean(best, k = 5, fill = NA, align = "right"))

model_system <- list(
  build_model("deterministic", formula = gdppc ~ I(abs(lag(gdppc) * (1 + gdppc_grwt)))),
  build_model("deterministic", formula = gdp ~ I(abs(gdppc * population))),
  build_model("parametric_distribution", formula = ~gdppc_grwt, distribution = "norm"),
  build_model("linear", formula = c1, boot = "resid"),
  build_model("univariate_fable",
              formula = v2x_polyarchy ~ error("A") + trend("N") + season("N"),
              method = "ets"),
  build_model("exogen", formula = ~psecprop),
  build_model("exogen", formula = ~population)
)
```

### Validation

`setup_simulator()` runs `validate_panel()` and `validate_system_closure()`
on your inputs before fitting anything. These check that the panel is
balanced, time is contiguous and integer-valued, the initial state at
`test_start - 1` has no NAs in any modelled outcome, and every variable
referenced by a formula is either modelled or supplied as a column. See
`?validate_panel` and `?validate_system_closure` if you want to run them
ad-hoc on a candidate panel.

```{r setup}
simulator_setup <- setup_simulator(
  models      = model_system,
  data        = df,
  train_start = 1970,
  test_start  = 2010,
  horizon     = 12,
  groupvar    = "gwcode",
  timevar     = "year",
  inner_sims  = 2,
  min_window  = 10
)
```

### Parallel execution and progress

The simulator no longer manages a `future` plan internally. Set one
yourself before calling `simulate_endogenr()`, and wrap the call in
`progressr::with_progress()` if you want a progress bar:

```{r simulate}
future::plan(future::multisession, workers = 2)

set.seed(42)
progressr::with_progress({
  res <- simulate_endogenr(nsim = 2, simulator_setup = simulator_setup)
})

future::plan(future::sequential)
```

### Scoring and plotting

`simulate_endogenr()` stamps `panel_unit` / `panel_time` attributes on the
result so `get_accuracy()` and `plotsim()` can infer the panel context
without any further configuration. Filter to the forecast window using a
plain `data.table` predicate.

```{r postprocess}
res <- res[res$year >= simulator_setup$test_start, ]

acc <- get_accuracy(res, "gdppc_grwt", df)
acc |>
  dplyr::summarize(dplyr::across(crps:winkler, ~ mean(.x))) |>
  dplyr::arrange(crps) |>
  knitr::kable()

plotsim(res, "gdppc", c(2, 20, 530), df)
plotsim(res, "gdppc_grwt", c(2, 20, 530), df)
plotsim(res, "v2x_polyarchy", c(2, 20, 530), df)
```

## Spatial lag

`endogenr` supports spatial-lag variables in the simulation. The setup is
two steps: (1) compute the spatial lag for the historical data, and
(2) register a `spatial_lag` model in the system so the lag is recomputed
at each simulated time step.

### Step 1: prepare spatial weights and the historical lag

`st_weights_from_sf()` builds a neighbourhood list and weights from an `sf`
object. The default is queen contiguity. For other schemes, call `sfdep`
directly and supply `nb`, `wt`, and `unit_ids` yourself (in the order they
appear in the spatial object).

Computing the historical lag can be fiddly when the panel is unbalanced or
has missing observations; the example below filters to units present in
the neighbourhood structure before computing the lag.

```{r spatial-lag, eval=FALSE}
library(endogenr)
library(dplyr)

df <- endogenr::example_data

# Load a map and filter to units present in the data
map <- poldat::cshp_gw_modifications(france_overseas = FALSE) |>
  dplyr::filter(end == as.Date("2019-12-31"))
map <- map |> dplyr::filter(gwcode %in% unique(df$gwcode))

# Build spatial weights (queen contiguity by default)
sf::sf_use_s2(FALSE)
neigh <- st_weights_from_sf(map, "gwcode", weights_args = list(allow_zero = TRUE))

# Compute the spatial lag of `best` for each year in the historical data
df <- df |>
  dplyr::filter(gwcode %in% neigh$unit_ids) |>
  dplyr::group_by(year) |>
  dplyr::mutate(
    sl_best = sfdep::st_lag(best, neigh$nb, neigh$wt, allow_zero = TRUE)
  ) |>
  dplyr::ungroup()
```

### Step 2: add a `spatial_lag` model to the system

A `spatial_lag` model is a cross-sectional transformation applied at every
simulated `t`. Reference the lag in other formulas as `lag(sl_y)` — using
the same-period value `sl_y` would create a circular dependency.

```{r spatial-lag-system, eval=FALSE}
c1 <- best ~ lag(best) + lag(sl_best) + lag(log(gdppc)) +
  lag(log(population)) + lag(psecprop) + lag(v2x_polyarchy) + lag(gdppc_grwt) +
  lag(zoo::rollmean(best, k = 5, fill = NA, align = "right"))

model_system <- list(
  # spatial_lag recomputes sl_best from `best` at each simulated t
  build_model("spatial_lag", formula = sl_best ~ best,
              nb = neigh$nb, wt = neigh$wt, unit_ids = neigh$unit_ids,
              island_default = 0),
  build_model("deterministic", formula = gdppc ~ I(abs(lag(gdppc) * (1 + gdppc_grwt)))),
  build_model("deterministic", formula = gdp ~ I(abs(gdppc * population))),
  build_model("parametric_distribution", formula = ~gdppc_grwt, distribution = "norm"),
  build_model("linear", formula = c1, boot = "resid"),
  build_model("univariate_fable",
              formula = v2x_polyarchy ~ error("A") + trend("N") + season("N"),
              method = "ets"),
  build_model("exogen", formula = ~psecprop),
  build_model("exogen", formula = ~population)
)

simulator_setup <- setup_simulator(
  models      = model_system,
  data        = df,
  train_start = 1970,
  test_start  = 2010,
  horizon     = 12,
  groupvar    = "gwcode",
  timevar     = "year",
  inner_sims  = 2,
  min_window  = 10
)

future::plan(future::multisession, workers = 2)
set.seed(42)
progressr::with_progress({
  res <- simulate_endogenr(nsim = 2, simulator_setup = simulator_setup)
})
future::plan(future::sequential)
```

## Long-horizon comparison

For each horizon `h`, the long-horizon API fits a single regression of
`y_{t+h}` on baseline covariates at `t`. Use it when you want a sanity
check against the dynamic simulator. See `?cv_long_horizon` for the
cross-validated entry point.

```{r long-horizon, eval=FALSE}
formulas <- list(
  lh_linear = gdppc_grwt ~ lag(gdppc_grwt) + lag(log(gdppc)) + lag(best)
)

lh_setup <- setup_long_horizon(
  data       = df,
  formulas   = formulas,
  horizons   = 1:12,
  groupvar   = "gwcode",
  timevar    = "year",
  train_end  = 2010
)

lh_forecasts <- forecast_long_horizon(
  lh_setup,
  data       = df,
  test_start = 2010,
  nsim       = 100,
  inner_sims = 10
)

lh_acc <- get_lh_accuracy(lh_forecasts, df, "gdppc_grwt", "gwcode", "year")
sim_acc <- get_accuracy(res, "gdppc_grwt", df)  # aggregate by horizon if comparing
compare_approaches(lh_acc, sim_acc)
```
