---
title: "Get started with endogenr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Get started with endogenr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`endogenr` builds probabilistic forecasts for systems of coupled statistical
models — where each model's predictions feed as inputs to the others.
This article walks through the full pipeline on a minimal three-equation system.

## The panel contract

`endogenr` expects a plain `data.frame` (or `data.table`; no `tsibble` needed)
with two identifier columns:

- **`groupvar`** — cross-sectional unit (e.g. country code).
- **`timevar`** — integer time index, contiguous *within* each unit.

Panels may be **unbalanced**: units may enter late or exit early.
`setup_system()` detects units that have no row at `test_start - 1` (the
forecast origin), keeps their data for training but excludes them from the
simulation grid, and prints a message listing them.

## Three stages

| Stage | Function | What it does |
|---|---|---|
| 1 | `setup_system()` | Validates panel + system closure, builds the dependency DAG, computes topological execution order, materialises the simulation grid — *no fitting* |
| 2 | `fit_system(nsim = )` | Fits each model spec and stores `nsim` coefficient draws |
| 3 | `simulate_system()` | Predict-only dynamic loop — derives `nsim` from stored draws; returns **forecast-only** rows (`time >= test_start`) |

**Two sources of uncertainty:**

- `nsim` (set on `fit_system`) — *parameter / model uncertainty*: each draw uses
  a potentially different set of coefficients.
- `inner_sims` (set on `setup_system`) — *innovation uncertainty*: residual or
  distributional draws within each coefficient draw.

The ensemble has `nsim × inner_sims` trajectories per unit.

## A minimal closed system

```{r data-prep}
library(endogenr)
set.seed(1)
future::plan(future::sequential)

vars <- c("gdppc", "gdppc_grwt", "psecprop")
yrs  <- 1990:2014
df   <- as.data.frame(endogenr::example_data)
df   <- df[df$year %in% yrs, c("gwcode", "year", vars)]

good <- function(d) {
  nrow(d) == length(yrs) &&
    all(vapply(d[vars], function(x) all(is.finite(x)), logical(1)))
}
keep <- names(which(vapply(split(df, df$gwcode), good, logical(1))))
keep <- head(keep, 6L)
df   <- df[df$gwcode %in% as.integer(keep), ]
```

The system below has three equations:

1. **`gdppc_grwt`** (linear OLS with residual bootstrap) — depends on lagged
   log GDP per capita and lagged secondary-school enrolment.
2. **`psecprop`** (exogenous) — observed values are copied from `df` for the
   whole horizon; no model is fitted.
3. **`gdppc`** (deterministic) — `gdppc` is updated each step from the previous
   `gdppc` and the freshly simulated `gdppc_grwt`.

Every right-hand-side variable is either produced by another model in the system
or supplied by an `exogen` — the system is **closed**.
The `exogen` requires that observed values exist for every unit through
`test_start + horizon - 1 = 2014`.

```{r run-pipeline}
model_system <- list(
  build_model("linear",
              formula = gdppc_grwt ~ lag(log(gdppc)) + lag(psecprop),
              boot = "resid"),
  build_model("exogen", formula = ~psecprop),
  build_model("deterministic",
              formula = gdppc ~ I(abs(lag(gdppc) * (1 + gdppc_grwt))))
)

sys <- setup_system(model_system, df,
                    train_start = 1990, test_start = 2010, horizon = 5,
                    groupvar = "gwcode", timevar = "year", inner_sims = 2)
fit <- fit_system(sys, nsim = 2)
res <- simulate_system(fit)
head(res)
```

## Automatic validation

`setup_system()` runs two validators before any fitting occurs:

- **`validate_panel()`** checks: integer time, contiguous within each unit,
  and a complete initial state at `test_start - 1` for every origin unit.
- **`validate_system_closure()`** checks: every RHS variable is produced by
  another model or present as a data column, no two models share an outcome,
  and all formula-referenced columns exist in `data`.

Units with no row at `test_start - 1` are printed by `setup_system()` and
excluded from the simulation grid; they still contribute to training.

## Score and plot

`simulate_system()` stamps `panel_unit` / `panel_time` attributes on the result,
so the post-processing helpers infer the panel context automatically — no need
to pass `groupvar`/`timevar` again.

```{r score-and-plot}
acc <- get_accuracy(res, "gdppc_grwt", df, test_start = 2010, by = "horizon")
acc
plotsim(res, "gdppc", head(keep, 3L), df)
```

`get_accuracy()` returns mean CRPS, MAE, and Winkler interval score
(default coverage level 50 %) per grouping. Pass `by = "horizon"` to score
per forecast step. `sim_to_dist()` converts the ensemble to summary quantiles
for custom summaries.

## Parallelism

`endogenr` does not manage a `future` plan — set one before calling and
wrap in `progressr::with_progress()` for a progress bar:

```{r parallel-note, eval = FALSE}
future::plan(future::multisession, workers = 4)
progressr::with_progress({
  res <- simulate_system(fit)
})
future::plan(future::sequential)
```

Leave the plan as `future::sequential` (as this vignette does) for
single-process, deterministic builds.
