---
title: "Experiment grids and the long-horizon benchmark"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Experiment grids and the long-horizon benchmark}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r data-prep}
library(endogenr)
set.seed(1)
future::plan(future::sequential)

vars <- c("gdppc", "gdppc_grwt", "psecprop")
yrs  <- 1990:2014
df   <- as.data.frame(endogenr::example_data)
df   <- df[df$year %in% yrs, c("gwcode", "year", vars)]

good <- function(d) {
  nrow(d) == length(yrs) &&
    all(vapply(d[vars], function(x) all(is.finite(x)), logical(1)))
}
keep <- names(which(vapply(split(df, df$gwcode), good, logical(1))))
keep <- head(keep, 6L)
df   <- df[df$gwcode %in% as.integer(keep), ]
```

## Experiment grid

`run_experiments()` runs the full `setup_system() → fit_system() →
simulate_system()` pipeline across a Cartesian product of:

- model systems (`models`)
- `train_start` (one or more values)
- `test_start` (one or more values)
- optionally `windows` (a named list of `window_config()` objects — adds a
  fourth axis)

Results are stacked into one `data.table` stamped with `.experiment`, `model`,
`train_start`, `test_start`, and `window_cfg`.

`vary_model(base, variants)` builds the named list of systems by swapping
specs into a base system **by outcome variable** — it never appends new specs,
only replaces existing ones with the same outcome.

`get_experiment_accuracy(results, outcome, truth)` scores each experiment
against its own `test_start`.

```{r experiments}
base <- list(
  build_model("linear",
              formula = gdppc_grwt ~ lag(log(gdppc)),
              boot = "resid"),
  build_model("exogen", formula = ~psecprop),
  build_model("deterministic",
              formula = gdppc ~ I(abs(lag(gdppc) * (1 + gdppc_grwt))))
)

# Swap the gdppc_grwt spec between two formula variants
systems <- vary_model(base, list(
  plus_edu = build_model("linear",
                         formula = gdppc_grwt ~ lag(log(gdppc)) + lag(psecprop),
                         boot = "resid")
))
names(systems)

exp_res <- run_experiments(
  data = df, models = systems,
  train_start = 1990, test_start = 2010, horizon = 5,
  groupvar = "gwcode", timevar = "year",
  inner_sims = 2, nsim = 2
)

get_experiment_accuracy(exp_res, "gdppc", df)
```

`systems` contains the base system (named `"model1"` automatically by
`vary_model`) and the `"plus_edu"` variant.  `get_experiment_accuracy` returns
one row per `.experiment` with mean CRPS, MAE, and Winkler scores.

**Parallelism axes:**  set `parallel = "experiments"` (default; maps the
experiment grid in parallel, each experiment sequential) or
`parallel = "draws"` (experiments sequential, each experiment's
`fit`/`simulate` parallelises).  Use `window_config()` to add a window axis:

```{r window-config, eval = FALSE}
run_experiments(
  data = df, models = systems,
  train_start = 1990, test_start = 2010, horizon = 5,
  groupvar = "gwcode", timevar = "year",
  inner_sims = 2, nsim = 2,
  windows = list(
    rand = window_config(),
    roll = window_config("rolling", width = 15, window_policy = "latest")
  )
)
```

## Long-horizon reduced-form benchmark

The long-horizon (LH) approach is *distinct* from the dynamic simulator: it
fits one **pooled** model per `(formula, horizon)` that leads the target
directly via `lead_horizon(y) ~ x`.  No dynamic loop is needed.

**When to use it:** as a benchmark.  An LH model is simpler and faster;
the dynamic simulator wins when feedback and cross-equation dynamics matter.
`compare_approaches()` stacks LH and simulation accuracy for side-by-side
comparison.

### `lead_horizon(y, h)` on the LHS

Inside a long-horizon formula, wrap the LHS outcome in `lead_horizon()`.
`setup_long_horizon()` substitutes the horizon `h` internally for each
`horizons` element — you do not supply `h` in the formula.  The RHS is
evaluated at the forecast origin (the last observed year), so write `lag()`
only for older history.

### Fit, forecast, and score

```{r long-horizon}
lh <- setup_long_horizon(
  data = df,
  formulas = list("y|y" = lead_horizon(log(gdppc)) ~ factor(gwcode) + log(gdppc)),
  horizons = 1:5,
  groupvar = "gwcode", timevar = "year",
  test_start = 2010, boot = "resid"
)

lh_res <- forecast_long_horizon(lh, data = df, nsim = 2, inner_sims = 5)
lh_acc <- get_lh_accuracy(lh_res, df, scale = "native", inverse = exp)
head(lh_acc)
```

**Scale controls:**

- `scale = "native"` (default) scores on the original outcome scale; supply
  `inverse = exp` when the LHS is `log(gdppc)` to back-transform the draws.
- `scale = "model"` scores on the modelled (log) scale; no `inverse` needed.

### Other long-horizon utilities

- `cv_long_horizon(data, formulas, horizons, groupvar, timevar, test_starts, ...)`
  — cross-validate across multiple `test_start` values and average scores over
  folds.
- `compare_approaches(lh_accuracy, sim_accuracy)` — stack LH and dynamic-sim
  accuracy into one table with an `approach` column.  Score the simulation with
  `get_accuracy(..., test_start = 2010, by = c("gwcode", "horizon"))` for a
  like-for-like comparison on the `(unit, horizon)` key.
