---
title: "Fixed effects, Mundlak means, and scenario parameters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fixed effects, Mundlak means, and scenario parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r data-prep}
library(endogenr)
set.seed(1)
future::plan(future::sequential)

vars <- c("gdppc", "gdppc_grwt", "psecprop")
yrs  <- 1990:2014
df   <- as.data.frame(endogenr::example_data)
df   <- df[df$year %in% yrs, c("gwcode", "year", vars)]

good <- function(d) {
  nrow(d) == length(yrs) &&
    all(vapply(d[vars], function(x) all(is.finite(x)), logical(1)))
}
keep <- names(which(vapply(split(df, df$gwcode), good, logical(1))))
keep <- head(keep, 6L)
df   <- df[df$gwcode %in% as.integer(keep), ]
```

## Unit fixed effects

`factor(gwcode)` in any pooled formula works as a standard dummy-variable
unit fixed effect.  The panel key column is present in every per-group frame,
so:

```r
gdppc_grwt ~ factor(gwcode) + lag(log(gdppc))
```

estimates a unit-specific intercept for every group in the training data.
The panel key must be referenced as `factor(<groupvar>)`.

## Time fixed effects and the forecast problem

`factor(year)` introduces **forecast years as unseen factor levels** — standard
`predict.lm()` would error.  `linear` models handle this automatically:

1. At **fit time**, `.detect_time_fe()` records the estimated year effects.
2. At **predict time**, the time column is neutralised to the baseline training
   year (so the factor contrast contributes 0) and a **scenario-drawn
   time-FE offset** is added instead — drawn from the estimated year effects.

**Constraint (enforced with a clear error):** the `timevar` may enter the
formula *only* through `factor(timevar)`.  Combining `factor(gwcode)` and
`factor(year)` gives a genuine **two-way fixed-effects (TWFE)** model that
simulates forward.

## TWFE example

```{r twfe}
twfe <- list(
  build_model("linear",
              formula = gdppc_grwt ~ factor(gwcode) + factor(year) +
                lag(log(gdppc)) + lag(psecprop),
              boot = "resid"),
  build_model("exogen", formula = ~psecprop),
  build_model("deterministic",
              formula = gdppc ~ I(abs(lag(gdppc) * (1 + gdppc_grwt))))
)

sys <- setup_system(twfe, df,
                    train_start = 1990, test_start = 2010, horizon = 5,
                    groupvar = "gwcode", timevar = "year", inner_sims = 2)
fit <- fit_system(sys, nsim = 2)
res <- simulate_system(fit)
```

## Inspect and configure scenario parameters

`setup_param(fit)` returns an `endogenr_scenario_params` object keyed by
outcome variable, covering **every** model in the fitted system.  All
stochastic draws (time-FE offsets) are **baked at build time** so you can
inspect the exact values that will enter the run.

Each entry contains:

| Field | Description |
|---|---|
| `type` | Model type string (e.g. `"linear"`) |
| `time_fe` | `endogenr_fe_param` or `NULL`.  Fields: `kind`, `term`, `var`, `ref`, `levels`, `effects`, `effects_by_draw`, `dims`, `active`, `heuristic`, `values` (baked array `nsim × inner_sims × horizon`). |
| `unit_fe` | `endogenr_fe_param` or `NULL`.  Default `active = FALSE` (persist historical effects unchanged). |
| `coef` | `endogenr_coef_param` or `NULL`.  Fields: `estimates`, `beta_by_draw`, `linear`, `overrides`, `effective`. |

The object carries a `dims` attribute:
`list(nsim, inner_sims, horizon, test_start, units, timevar, unitvar)`.

```{r setup-param}
set.seed(2)
sp <- setup_param(fit)
sp
```

The `time_fe` block for `gdppc_grwt` confirms `factor(year)` detection and
shows the distribution of estimated year effects plus a summary of the baked
offset array.

## Time-FE heuristic helpers

Replace `sp[["outcome"]]$time_fe` with the output of one of the helper
functions to re-bake the draws under a different strategy before passing `sp`
to `simulate_system()`.

- **`fe_resample(block)`** (default) — sample from estimated year effects per
  `(draw, inner_sim, step)`.
- **`fe_distribution(block, sd = ..)`** — Normal draws; optionally fix `mean`
  and `sd` (filled from effects per draw when omitted).
- **`fe_fixed(block, value = 0)`** — constant offset for all trajectories.
- **`fe_converge(block, to, path)`** — path-shaped convergence toward a target
  (`"zero"`, `"mean"`, numeric, or character unit ids).

All helpers accept `to`, `factor`, `path` (`"constant"` / `"linear"` /
`"sigmoid"`), `mid`, `steep` arguments for fine-grained control.  They
consume RNG at call time — `set.seed()` before the helper for reproducibility.

```{r tfe-distribution}
set.seed(2)
sp <- setup_param(fit)
sp[["gdppc_grwt"]]$time_fe <- fe_distribution(sp[["gdppc_grwt"]]$time_fe, sd = 0.01)
res_dist <- simulate_system(fit, scenario_params = sp)
nrow(res_dist)
```

Converge the time-FE offset toward zero over the horizon:

```{r tfe-converge, eval = FALSE}
set.seed(2); sp <- setup_param(fit)
sp[["gdppc_grwt"]]$time_fe <- fe_converge(sp[["gdppc_grwt"]]$time_fe,
                                           to = 0, path = "linear")
res_conv <- simulate_system(fit, scenario_params = sp)
```

## Unit-FE heuristic helpers

`factor(unit)` models store a `unit_fe` block with `active = FALSE` by
default (historical unit effects stay in the design matrix unchanged).
Apply `fe_converge` or `fe_persist` (with non-default `to`/`path`) to
activate explicit scenario control:

```{r ufe-converge, eval = FALSE}
# Converge unit-FE offsets to zero over the horizon (e.g. neutralise
# country-specific intercepts by the end of the forecast window)
set.seed(2); sp <- setup_param(fit)
sp[["gdppc_grwt"]]$unit_fe <- fe_converge(sp[["gdppc_grwt"]]$unit_fe,
                                           to = 0, path = "linear")
res_ufe <- simulate_system(fit, scenario_params = sp)
```

## Coefficient overrides (dynamic simulation parameters)

`coef_override(block, term, value)` bakes overrides into the coef block.
`value` may be:

- A **scalar** — constant across all draws and forecast steps.
- A **length-`horizon` numeric vector** — per-step, same across draws.
- A **`function(h, beta_hat)`** — returns the override β* at step `h` given
  the per-draw fitted estimate.

The term name must be in `colnames(sp[[oc]]$coef$beta_by_draw)` — the alias
names from `get_coefficients()` (e.g. `lag_log_gdppc`, `lag_psecprop`).
Overrides are only supported for `linear` models; `coef_override()` errors
immediately for other types.

`predict.linear` swaps only the targeted coefficient(s) while the residual
scale and prediction intervals are untouched, preserving forecast uncertainty.

```{r coef-override}
# Function override: halve the autoregressive persistence of lag_log_gdppc
set.seed(2); sp <- setup_param(fit)
sp[["gdppc_grwt"]]$coef <- coef_override(
  sp[["gdppc_grwt"]]$coef, "lag_log_gdppc",
  function(h, beta_hat) beta_hat * 0.5
)
res_ovr <- simulate_system(fit, scenario_params = sp)
nrow(res_ovr)
```

Scalar example: pin the `lag_psecprop` coefficient to zero (testing exclusion
of education):

```{r coef-scalar, eval = FALSE}
sp[["gdppc_grwt"]]$coef <- coef_override(sp[["gdppc_grwt"]]$coef,
                                          "lag_psecprop", 0)
res_no_edu <- simulate_system(fit, scenario_params = sp)
```

Pass `value = NULL` to clear a previously applied override:

```{r coef-clear, eval = FALSE}
sp[["gdppc_grwt"]]$coef <- coef_override(sp[["gdppc_grwt"]]$coef,
                                          "lag_psecprop", NULL)
```

## Mundlak between-effects

The Mundlak (1978) correction decomposes a time-varying covariate into its
within-unit deviation and its unit mean, adding the mean as a separate
predictor to absorb between-unit confounding.  `mundlak_means(vars)` builds a
list of `deterministic` specs that maintain running within-unit means (`m_*`
columns) via `hist_mean()`, updated at every simulation step.

**Conventions enforced by `validate_panel()`:**

- Each `m_*` output column must already exist in `data` before `setup_system()`.
  For units present at `test_start - 1`, the column must be **non-NA** there
  (it is the initial state the simulation bootstraps from).  Pre-compute the
  running mean over the training window and leave the forecast rows `NA` — the
  deterministic model updates them each step.
- Each source column must be model-produced or a panel key.

```{r mundlak}
# Pre-compute the expanding within-unit mean for the training period.
# The value at year 2009 (test_start - 1) is the required initial state.
df <- df[order(df$gwcode, df$year), ]
df$m_psecprop <- unlist(
  lapply(split(df$psecprop, df$gwcode), endogenr::hist_mean)
)
df$m_psecprop[df$year >= 2010] <- NA_real_  # forecast rows filled by simulation

mund <- c(
  list(build_model("exogen", formula = ~psecprop)),
  mundlak_means("psecprop"),        # -> m_psecprop = hist_mean(psecprop)
  list(
    build_model("linear",
                formula = gdppc_grwt ~ lag(log(gdppc)) + psecprop + m_psecprop,
                boot = "resid"),
    build_model("deterministic",
                formula = gdppc ~ I(abs(lag(gdppc) * (1 + gdppc_grwt))))
  )
)

sysm <- setup_system(mund, df,
                     train_start = 1990, test_start = 2010, horizon = 5,
                     groupvar = "gwcode", timevar = "year", inner_sims = 2)
fitm <- fit_system(sysm, nsim = 2)
resm <- simulate_system(fitm)
nrow(resm)
```

`psecprop` is exogenous, so `m_psecprop = hist_mean(psecprop)` introduces no
dependency cycle.  `window = Inf` (the default) gives an expanding
(cumulative) mean; a positive integer gives a trailing window.  Per-variable
windows can be set with a named vector: `mundlak_means(c("x", "z"),
window = c(m_x = 5, m_z = Inf))`.
