---
title: "Fixed effects, Mundlak means, and scenario parameters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fixed effects, Mundlak means, and scenario parameters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r data-prep}
library(endogenr)
set.seed(1)
future::plan(future::sequential)

vars <- c("gdppc", "gdppc_grwt", "psecprop")
yrs  <- 1990:2014
df   <- as.data.frame(endogenr::example_data)
df   <- df[df$year %in% yrs, c("gwcode", "year", vars)]

good <- function(d) {
  nrow(d) == length(yrs) &&
    all(vapply(d[vars], function(x) all(is.finite(x)), logical(1)))
}
keep <- names(which(vapply(split(df, df$gwcode), good, logical(1))))
keep <- head(keep, 6L)
df   <- df[df$gwcode %in% as.integer(keep), ]
```

## Unit fixed effects

`factor(gwcode)` in any pooled formula works as a standard dummy-variable
unit fixed effect.  The panel key column is present in every per-group frame,
so:

```r
gdppc_grwt ~ factor(gwcode) + lag(log(gdppc))
```

estimates a unit-specific intercept for every group in the training data.
The panel key must be referenced as `factor(<groupvar>)`.

## Time fixed effects and the forecast problem

`factor(year)` introduces **forecast years as unseen factor levels** — standard
`predict.lm()` would error.  `linear` models handle this automatically:

1. At **fit time**, `.detect_time_fe()` records the estimated year effects.
2. At **predict time**, the time column is neutralised to the baseline training
   year (so the factor contrast contributes 0) and a **scenario-drawn
   time-FE offset** is added instead — drawn from the estimated year effects.

**Constraint (enforced with a clear error):** the `timevar` may enter the
formula *only* through `factor(timevar)`.  Combining `factor(gwcode)` and
`factor(year)` gives a genuine **two-way fixed-effects (TWFE)** model that
simulates forward.

## TWFE example

```{r twfe}
twfe <- list(
  build_model("linear",
              formula = gdppc_grwt ~ factor(gwcode) + factor(year) +
                lag(log(gdppc)) + lag(psecprop),
              boot = "resid"),
  build_model("exogen", formula = ~psecprop),
  build_model("deterministic",
              formula = gdppc ~ I(abs(lag(gdppc) * (1 + gdppc_grwt))))
)

sys <- setup_system(twfe, df,
                    train_start = 1990, test_start = 2010, horizon = 5,
                    groupvar = "gwcode", timevar = "year", inner_sims = 2)
fit <- fit_system(sys, nsim = 2)
res <- simulate_system(fit)
```

## Inspect and configure scenario parameters

`setup_param(fit)` returns an editable `endogenr_scenario_params` list keyed
by outcome variable.  Each entry contains:

| Field | Description |
|---|---|
| `type` | Model type string (e.g. `"linear"`) |
| `time_fe` | `NULL`, or a list with `term`, `timevar`, `n_levels`, `effects` (numeric vector, baseline = 0), `summary` (mean/sd/min/max) |
| `policy` | Time-FE draw policy: `$type` is `"resample"` (default), `"distribution"`, or `"fixed"` |
| `coef_table` | `data.frame` of non-FE coefficient estimates and standard errors, or `NULL` |
| `estimates` | Named numeric vector of overridable coefficient estimates |
| `coefficients` | Empty `list()` for user coefficient overrides |

```{r setup-param}
sp <- setup_param(fit)
sp
```

The `time FE` block for `gdppc_grwt` confirms that `factor(year)` was detected,
and shows the distribution of the estimated year effects.

## Time-FE policies

`sp[["gdppc_grwt"]]$policy$type` controls how a year-effect offset is drawn
for each simulated trajectory:

- **`"resample"`** (default) — sample uniformly from estimated year effects.
- **`"distribution"`** — Normal draw; optionally set `$mean` and `$sd` (filled
  from the effects if omitted).
- **`"fixed"`** — constant `$value` for all trajectories.

One offset is drawn per trajectory and shared across all rows of that
trajectory.

```{r policy-override}
sp[["gdppc_grwt"]]$policy <- list(type = "distribution", sd = 0.01)
res_dist <- simulate_system(fit, scenario_params = sp)
nrow(res_dist)
```

## Coefficient overrides (dynamic simulation parameters)

`sp[["outcome"]]$coefficients[["<term>"]]` accepts:

- A **scalar** — constant across all forecast steps.
- A **length-`horizon` numeric vector** — per-step trajectory.
- A **`function(h, beta_hat)`** — returns the override β* at step `h` given
  the per-draw fitted estimate.

The term name must be one of `names(sp[[oc]]$estimates)` — the alias names
from `get_coefficients()` (e.g. `lag_log_gdppc`, `lag_psecprop`).  Overrides
are validated in `simulate_system()` and are **`linear`-only**
(an error is thrown for other model types).

`predict.linear` swaps only the targeted coefficient(s) while the residual
scale and prediction intervals are untouched, preserving forecast uncertainty.

```{r coef-override}
# Function override: halve the autoregressive persistence of lag_log_gdppc
sp[["gdppc_grwt"]]$coefficients[["lag_log_gdppc"]] <-
  function(h, beta_hat) beta_hat * 0.5

res_ovr <- simulate_system(fit, scenario_params = sp)
nrow(res_ovr)
```

Scalar example: pin the `lag_psecprop` coefficient to zero (testing exclusion
of education):

```{r coef-scalar, eval = FALSE}
sp[["gdppc_grwt"]]$coefficients[["lag_psecprop"]] <- 0
res_no_edu <- simulate_system(fit, scenario_params = sp)
```

## Mundlak between-effects

The Mundlak (1978) correction decomposes a time-varying covariate into its
within-unit deviation and its unit mean, adding the mean as a separate
predictor to absorb between-unit confounding.  `mundlak_means(vars)` builds a
list of `deterministic` specs that maintain running within-unit means (`m_*`
columns) via `hist_mean()`, updated at every simulation step.

**Conventions enforced by `validate_panel()`:**

- Each `m_*` output column must already exist in `data` before `setup_system()`.
  For units present at `test_start - 1`, the column must be **non-NA** there
  (it is the initial state the simulation bootstraps from).  Pre-compute the
  running mean over the training window and leave the forecast rows `NA` — the
  deterministic model updates them each step.
- Each source column must be model-produced or a panel key.

```{r mundlak}
# Pre-compute the expanding within-unit mean for the training period.
# The value at year 2009 (test_start - 1) is the required initial state.
df <- df[order(df$gwcode, df$year), ]
df$m_psecprop <- unlist(
  lapply(split(df$psecprop, df$gwcode), endogenr::hist_mean)
)
df$m_psecprop[df$year >= 2010] <- NA_real_  # forecast rows filled by simulation

mund <- c(
  list(build_model("exogen", formula = ~psecprop)),
  mundlak_means("psecprop"),        # -> m_psecprop = hist_mean(psecprop)
  list(
    build_model("linear",
                formula = gdppc_grwt ~ lag(log(gdppc)) + psecprop + m_psecprop,
                boot = "resid"),
    build_model("deterministic",
                formula = gdppc ~ I(abs(lag(gdppc) * (1 + gdppc_grwt))))
  )
)

sysm <- setup_system(mund, df,
                     train_start = 1990, test_start = 2010, horizon = 5,
                     groupvar = "gwcode", timevar = "year", inner_sims = 2)
fitm <- fit_system(sysm, nsim = 2)
resm <- simulate_system(fitm)
nrow(resm)
```

`psecprop` is exogenous, so `m_psecprop = hist_mean(psecprop)` introduces no
dependency cycle.  `window = Inf` (the default) gives an expanding
(cumulative) mean; a positive integer gives a trailing window.  Per-variable
windows can be set with a named vector: `mundlak_means(c("x", "z"),
window = c(m_x = 5, m_z = Inf))`.
