---
title: "Formulas and panel conventions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Formulas and panel conventions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`endogenr` uses standard R formula syntax but evaluates formulas in two
stages so that within-unit time-series operations and pooled design-matrix
operations never collide.

## Two-stage formula evaluation

**Stage 1 (panel-aware)** — `panel_materialize()` scans the formula for
*maximal time-series sub-expressions* (calls whose head is in the registered
set; see below) and evaluates each one *per unit* in time order, storing the
result in a synthetic `.pt#` column.  The formula is rewritten to reference
those columns.

**Stage 2 (pooled)** — the rewritten formula is fitted with a pooled `lm()`
or `glm()` on all units at once.  Design constructs resolved here —
`factor()`, `poly()`, `bs()`/`ns()`, interactions (`a:b`, `a*b`), `-1` —
are reproduced at predict time through `predict`'s stored `predvars`/
`xlevels`.  This is why `poly(x, 2)` works even when a single unit has few
rows: the polynomial basis is built across all units.

## The time-series function registry

These call heads are treated as within-unit and materialised in Stage 1
(matched bare or `pkg::`-qualified; the full call including its arguments is
materialised, so `lag(zoo::rollmean(x, 5))` is always within-unit):

| Family | Functions |
|---|---|
| Positional | `lag`, `lead`, `lead_horizon`, `shift`, `diff` |
| Rolling (zoo/data.table) | `rollmean`/`r`/`l`, `rollsum`/`r`/`l`, `rollmax`/`r`/`l`, `rollmedian`/`r`/`l`, `rollapply`/`r`/`l`, `frollmean`, `frollsum`, `frollapply`, `frollmax`, `frollmin` |
| Cumulative | `cumsum`, `cumprod`, `cummax`, `cummin`, `decay_since_event`, `time_since_event`, `intensity_decay`, `hist_mean` |

A call to any of these — however deeply nested — is materialised as a unit
block.  Anything *not* in this registry (e.g. a custom `my_smooth()` function)
is evaluated pooled, which is wrong across unit boundaries.

## Positional semantics of `lag` / `lead` / `shift` / `diff`

These are bound *uniformly* by `endogenr` regardless of which packages are
attached, so `data.table` or `dplyr` need not be loaded:

- `lag(x, n)` — positional forward-shift by `n` within unit (pad head with
  `NA`).  Type-preserving: a lagged factor keeps its class and levels.
- `lead(x, n)` — pulls ahead by `n` (pad tail with `NA`).
- `lag(x, -1)` **errors** — use `lead(x, 1)` instead.
- `diff(x)` — length-stable (pads leading rows with `NA` rather than
  shortening the vector).

## The pitfall and its fixes

A custom within-unit function used *unwrapped* — not in the registry — is
evaluated **pooled** (across unit boundaries, which is wrong).

Fix in order of preference:

1. **Wrap in `lag()`** — the whole call is then materialised as a unit
   block: `lag(my_smooth(x))`.
2. **Precompute** it as a data column before `setup_system()`.
3. **Register** via the `ts_fns` argument available on the long-horizon
   fitters; for the main pipeline, use options 1 or 2.

## Naming contract

After Stage 1, the `.pt#` columns are renamed to `janitor`-cleaned deparsed
expressions before the pooled fit, so coefficient names match base `lm()`:

| Formula term | Coefficient name |
|---|---|
| `lag(x)` | `lag_x` |
| `lag(log(gdppc))` | `lag_log_gdppc` |
| `lag(psecprop)` | `lag_psecprop` |

Internal `.pt#` names never surface in `get_coefficients()` output.
**These alias names are also what you use for coefficient overrides** in
scenario parameters (see `vignette("fixed-effects-and-scenarios")`).

```{r naming-demo}
library(endogenr)
set.seed(1)
future::plan(future::sequential)

vars <- c("gdppc", "gdppc_grwt", "psecprop")
yrs  <- 1990:2014
df   <- as.data.frame(endogenr::example_data)
df   <- df[df$year %in% yrs, c("gwcode", "year", vars)]

good <- function(d) {
  nrow(d) == length(yrs) &&
    all(vapply(d[vars], function(x) all(is.finite(x)), logical(1)))
}
keep <- names(which(vapply(split(df, df$gwcode), good, logical(1))))
keep <- head(keep, 6L)
df   <- df[df$gwcode %in% as.integer(keep), ]

model_system <- list(
  build_model("linear",
              formula = gdppc_grwt ~ lag(log(gdppc)) + lag(psecprop),
              boot = "resid"),
  build_model("exogen", formula = ~psecprop),
  build_model("deterministic",
              formula = gdppc ~ I(abs(lag(gdppc) * (1 + gdppc_grwt))))
)

sys <- setup_system(model_system, df,
                    train_start = 1990, test_start = 2010, horizon = 5,
                    groupvar = "gwcode", timevar = "year", inner_sims = 2)
fit <- fit_system(sys, nsim = 2)

# Term names in the fitted linear model
coefs <- get_coefficients(fit)
coefs[coefs$outcome == "gdppc_grwt", c("term", "estimate")]
```

The terms `lag_log_gdppc` and `lag_psecprop` match exactly what you would see
in a plain `lm()` call — and they are the keys you use for coefficient
overrides in `setup_param()`.

## Formula helper functions

### `hist_mean()` — trailing or expanding within-unit mean

`hist_mean(x, window = Inf, min_obs = 1L)` computes the mean of the last
`window` observations position by position.  `Inf` gives an expanding
(cumulative) mean.  Returns `NA_real_` until `min_obs` non-NA values are
available.

```{r hist-mean}
hist_mean(c(1, 2, 3, 4))              # expanding mean
hist_mean(c(1, 2, 3, 4), window = 2) # trailing 2-step window
```

Used inside a formula it is materialised per unit in time order, so
`hist_mean(psecprop)` is the cumulative within-unit mean of `psecprop` up to
each row.

### `decay_since_event()` and friends

```{r decay-demo}
# 1.0 at the event, decays toward exp(-0.5 * 50) thereafter
decay_since_event(c(0, 0, 1, 0, 0), lambda = 0.5)
```

- `decay_since_event(event, lambda, max_years = 50)` — exponential decay from
  the most recent event; peak = 1.
- `time_since_event(event, lambda, max_years = 50)` — inverse: 0 at event,
  rising toward 1.
- `intensity_decay(event, intensity, lambda, max_years = 50)` — same but peak
  equals the event intensity; multiple events accumulate.

All three consume the entire prior unit series (`.required_history()` returns
`Inf`), so they always use full per-unit history regardless of window settings.

## Output bounds

`build_model(..., bounds = c(lower, upper))` clamps that model's outcome to
`[lower, upper]` at every forecast step before it feeds later steps.
Useful for stabilising autoregressive feedback:

```{r bounds-example, eval = FALSE}
build_model("linear",
            formula = gdppc_grwt ~ lag(log(gdppc)),
            boot = "resid",
            bounds = c(-1, 1))   # clamp growth rate
```

`Inf` is allowed for a one-sided limit (`c(0, Inf)` for a non-negative
outcome).  A draw that remains non-finite after clamping is reset to a
finite in-range fallback (midpoint of two finite bounds, or the finite bound
itself), so a bounded outcome never propagates `NaN`/`Inf`.

## User functions on parallel workers

When a formula references a user-defined function, pass it to `setup_system()`
so it is forwarded to parallel workers in both `fit_system()` and
`simulate_system()`:

```{r globals-example, eval = FALSE}
my_transform <- function(x) log1p(pmax(x, 0))

sys <- setup_system(
  model_system, df,
  train_start = 1990, test_start = 2010, horizon = 5,
  groupvar = "gwcode", timevar = "year", inner_sims = 2,
  globals = list(my_transform = my_transform)
)
```
