---
title: "Model families and predictive draws"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model families and predictive draws}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`build_model(type, formula, ..., bounds = NULL)` returns a lightweight spec;
no data is needed and no fitting happens here.  Fitting is deferred to
`fit_system()`, which means the same spec list can be re-used across different
training windows, bootstrap draws, and experiment grids.

## The 10 model types

### Imports-only families (safe to run in all evaluated chunks)

**`"linear"`** — two-sided formula; pooled OLS.  Optional `boot = "resid"` or
`"wild"` for residual or wild bootstrap (introduces parameter variation across
`nsim` draws).

```{r linear-example}
library(endogenr)
build_model("linear",
            formula = gdppc_grwt ~ lag(log(gdppc)) + lag(psecprop),
            boot = "resid")
```

**`"glm"`** — two-sided formula; pooled GLM.  `family = stats::gaussian()`
by default; pass any `stats::family()`.  Optional `boot`.

```{r glm-example, eval = FALSE}
build_model("glm",
            formula = event ~ lag(log(gdppc)),
            family = stats::binomial())
```

**`"deterministic"`** — two-sided formula; RHS **must** be wrapped in `I()`.
Evaluated (not fitted) each step.  No `boot`, no extra arguments.

```{r det-example}
build_model("deterministic",
            formula = gdppc ~ I(abs(lag(gdppc) * (1 + gdppc_grwt))))
```

**`"exogen"`** — one-sided formula `~var`.  Copies observed values from `data`
into the simulation grid; the variable must be present for every unit through
`test_start + horizon - 1`.  No fitting.

```{r exogen-example}
build_model("exogen", formula = ~psecprop)
```

### Suggests-gated families (shown syntax-only; not evaluated)

These require packages listed under `Suggests` in `DESCRIPTION`.

**`"parametric_distribution"`** — requires `fitdistrplus`.  One-sided `~var`.
Fits a parametric distribution to the training window.  The bundled `"t_ls"`
(location-scale Student-t) derives starting values automatically.
`param_uncertainty = TRUE` propagates MLE uncertainty across draws.

```{r param-dist, eval = FALSE}
build_model("parametric_distribution",
            formula = ~gdppc_grwt,
            distribution = "t_ls",
            param_uncertainty = TRUE)
```

**`"univariate_fable"`** — requires `fable`/`fabletools`/`tsibble`.
Two-sided fable formula.

```{r fable, eval = FALSE}
build_model("univariate_fable",
            formula = gdppc_grwt ~ error("A") + trend("N") + season("N"),
            method = "ets")
```

**`"heterolm"`** — requires `heterolm`.  Two-sided mean formula plus a
one-sided log-variance formula (defaults to `~1`).

```{r heterolm, eval = FALSE}
build_model("heterolm",
            formula = gdppc_grwt ~ lag(log(gdppc)) + lag(psecprop),
            variance = ~lag(log(gdppc)))
```

**`"glmmTMB"`** — requires `glmmTMB`.  Two-sided formula with lme4-style
random-effects bars and glmmTMB covariance-structure wrappers.  Grouping
factors and coordinate columns used in covariance structures (e.g. `region`,
`times`) must be produced by some model — add an `exogen` to carry them
forward, or use a panel key.

```{r glmmTMB, eval = FALSE}
build_model("glmmTMB",
            formula = gdppc_grwt ~ lag(log(gdppc)) + (1 | gwcode),
            family = stats::gaussian())
```

**`"gamlss"`** — requires `gamlss`.  Two-sided `formula` for the `mu` (mean)
parameter; optional `sigma.formula`, `nu.formula`, `tau.formula`.  Grouping
factors inside `random()`/`ra()`/`re()` must be produced or be a panel key.

```{r gamlss, eval = FALSE}
build_model("gamlss",
            formula = gdppc_grwt ~ lag(log(gdppc)),
            sigma.formula = ~lag(log(gdppc)),
            family = gamlss.dist::NO())
```

**`"spatial_lag"`** — requires `sfdep`.  Two-sided formula; use `lag(y)` to
avoid a circular dependency on a same-period outcome.  Pass `nb`, `wt`, and
`unit_ids` from `st_weights_from_sf()`.

```{r spatial-lag, eval = FALSE}
weights <- st_weights_from_sf(sf_object, unit_col = "gwcode")
build_model("spatial_lag",
            formula = sl_gdppc ~ lag(gdppc),
            nb = weights$nb, wt = weights$wt,
            unit_ids = weights$unit_ids)
```

## Predictive draws — `draw_predictive()`

The exported generic `draw_predictive(model, newdata, n_param, n_innov, ...)`
returns an `nrow(newdata) × max(n_param, 1) * max(n_innov, 1)` matrix of
response-scale draws:

- `n_param` — parameter-uncertainty draws (`0` = condition on point estimates).
- `n_innov` — innovation draws per parameter block (`0` = conditional mean).

Methods are implemented for `linear`, `glm`, `glmmTMB`, `parametric_distribution`,
and `univariate_fable`.  `deterministic`, `exogen`, and `spatial_lag` have no
stochastic predictive distribution and error if called directly.

```{r draw-predictive, eval = FALSE}
set.seed(1)
future::plan(future::sequential)

vars <- c("gdppc", "gdppc_grwt", "psecprop")
yrs  <- 1990:2014
df   <- as.data.frame(endogenr::example_data)
df   <- df[df$year %in% yrs, c("gwcode", "year", vars)]

good <- function(d) {
  nrow(d) == length(yrs) &&
    all(vapply(d[vars], function(x) all(is.finite(x)), logical(1)))
}
keep <- names(which(vapply(split(df, df$gwcode), good, logical(1))))
keep <- head(keep, 6L)
df   <- df[df$gwcode %in% as.integer(keep), ]

sys <- setup_system(
  list(
    build_model("linear",
                formula = gdppc_grwt ~ lag(log(gdppc)) + lag(psecprop),
                boot = "resid"),
    build_model("exogen", formula = ~psecprop),
    build_model("deterministic",
                formula = gdppc ~ I(abs(lag(gdppc) * (1 + gdppc_grwt))))
  ),
  df, train_start = 1990, test_start = 2010, horizon = 5,
  groupvar = "gwcode", timevar = "year", inner_sims = 2
)
fit <- fit_system(sys, nsim = 2)

# draw_predictive() needs newdata with materialized columns (lag_log_gdppc,
# lag_psecprop); the simulation loop provides those automatically. For a direct
# call, use sim_to_dist() on simulate_system() output instead — it returns the
# same per-cell draw vectors as a list-column.
linear_model <- Filter(function(m) inherits(m, "linear"), fit$fitted_models)[[1]]
# n_param = 0: condition on point estimates; n_innov = 4: four residual draws
# draws <- draw_predictive(linear_model, prepared_newdata, n_param = 0L, n_innov = 4L)
```

## Ensemble semantics — what you must know

The full ensemble is `nsim × inner_sims` trajectories, but parameter
uncertainty enters **per family**:

- **`linear` / `glm` / `heterolm` / `glmmTMB`** gain across-draw parameter
  variation only when `min_window` (random-window refits) or `boot` is set;
  without either, every draw uses the same point estimates.
- **`parametric_distribution`** and **`univariate_fable`** are fitted *once
  and shared* across all `nsim` draws, so parameter uncertainty is zero —
  except `parametric_distribution` with `param_uncertainty = TRUE`.
- **`exogen`** carries no stochastic distribution at all; it copies observed
  values.

Innovations are drawn independently per `(unit, time, sim)` row.  The engine
does not capture cross-unit common shocks or cross-equation residual
correlation.  See `?endogenr` for the full capability roadmap.
