---
title: "Estimation windows and coefficient uncertainty"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimation windows and coefficient uncertainty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r data-prep}
library(endogenr)
set.seed(1)
future::plan(future::sequential)

vars <- c("gdppc", "gdppc_grwt", "psecprop")
yrs  <- 1990:2014
df   <- as.data.frame(endogenr::example_data)
df   <- df[df$year %in% yrs, c("gwcode", "year", vars)]

good <- function(d) {
  nrow(d) == length(yrs) &&
    all(vapply(d[vars], function(x) all(is.finite(x)), logical(1)))
}
keep <- names(which(vapply(split(df, df$gwcode), good, logical(1))))
keep <- head(keep, 6L)
df   <- df[df$gwcode %in% as.integer(keep), ]

model_system <- list(
  build_model("linear",
              formula = gdppc_grwt ~ lag(log(gdppc)) + lag(psecprop),
              boot = "resid"),
  build_model("exogen", formula = ~psecprop),
  build_model("deterministic",
              formula = gdppc ~ I(abs(lag(gdppc) * (1 + gdppc_grwt))))
)
```

## Where uncertainty enters

`endogenr` uses a nested Monte Carlo:

- **`nsim` draws** (set on `fit_system`) — parameter / model uncertainty.
  Each draw uses potentially different coefficients.
- **`inner_sims`** (set on `setup_system`) — innovation uncertainty.
  Each residual or distributional draw within a coefficient draw.

The ensemble has `nsim × inner_sims` trajectories per unit.

For the regression families (`linear`, `glm`, `heterolm`, `glmmTMB`),
across-draw parameter variation comes from `boot` (residual or wild bootstrap)
and/or random window refits (`min_window`).  Without either, all draws share
the same point estimates and the ensemble captures only innovation uncertainty.

## Random windows (`min_window`)

Setting `min_window` on `setup_system()` tells `fit_system()` to refit each
regression spec on a randomly drawn training window of at least `min_window`
time steps per coefficient draw (via `get_train_window()`).  Each draw can
therefore see a different amount of history and produce different coefficients.

`get_coefficients(fit)` surfaces the per-draw metadata:

| Column | Meaning |
|---|---|
| `.draw` | Draw index (1 … nsim) |
| `.window_start` | Start of the training window for that draw |
| `.window_end` | End of the training window for that draw |
| `outcome` | Model outcome variable |
| `term` | Coefficient alias name |
| `estimate` | Coefficient estimate |

```{r random-windows}
sys_mw <- setup_system(model_system, df,
                       train_start = 1990, test_start = 2010, horizon = 5,
                       groupvar = "gwcode", timevar = "year",
                       inner_sims = 2, min_window = 12)
fit_mw <- fit_system(sys_mw, nsim = 4)

unique(get_coefficients(fit_mw)[, c(".draw", ".window_start", ".window_end")])
```

Each draw uses a different window — the variation across draws is *model*
uncertainty (different training periods), on top of the residual bootstrap
variation from `boot = "resid"`.

**Without `min_window` and without `boot`**, all draws share the same fit
(`[train_start, test_start - 1]`) and reproduce a plain pooled `lm()` exactly.

## Sliding windows

`fit_system(sys, window = "rolling", width = 12)` or `window = "expanding"`
refits each regression spec on a *deterministic* grid of window ends (anchors
from the first valid window to `test_start - 1`, spaced by `step`).  There is
one fit per anchor; `simulate_system()` mixes the per-window fits according to
`window_policy`:

| Policy | Behaviour |
|---|---|
| `"latest"` (default) | Always use the most-recent window |
| `"equal"` | Each window with equal probability |
| `"decay"` | Geometric decay toward recent windows; tuned by `decay` |

Pass `weights` for full control: a `function(h, age)` or a
`horizon × n_window` numeric matrix.  `window_policy`/`weights` are ignored
with a warning for random/fixed-window fits.

```{r sliding-windows}
sys_base <- setup_system(model_system, df,
                         train_start = 1990, test_start = 2010, horizon = 5,
                         groupvar = "gwcode", timevar = "year",
                         inner_sims = 2, min_window = 12)

fit_roll <- fit_system(sys_base, nsim = 2, window = "rolling", width = 12)
res_roll <- simulate_system(fit_roll, window_policy = "equal")
nrow(res_roll)
```

`width` defaults to `min_window` (or `10` when that is unset).

## Coefficient-trajectory forecasting

`forecast_coefficients()` estimates a deterministic coefficient *path* by
refitting on an expanding or rolling grid of windows, then projects each
coefficient forward as a Gaussian random walk (optionally with drift).  The
forward distribution fans from the forecast origin's estimation uncertainty
(the origin fit's variance-covariance matrix) and the empirical evolution
covariance of the path.

```{r coef-forecast}
cf <- forecast_coefficients(fit_mw, window = "expanding")
head(cf[cf$.type == "forecast", c("outcome", "term", ".time", ".mean", ".q05", ".q95")])
```

`plot_coefficient_forecast(cf)` plots the observed coefficient path and the
forecast band, faceted by outcome and term:

```{r plot-coef-forecast}
plot_coefficient_forecast(cf)
```

Also available:

- `plot_coefficients(fit)` — per-term violin plots of the coefficient draws
  across the `nsim` fitted models.
- `plot_estimates(fit)` — posterior predictive ribbons of model-fitted values.

## Known caveats

Using `linear` + `boot` + `min_window` together can **triple-count** parameter
uncertainty: random window variation, residual bootstrap resampling, and the
`predict.lm(se.fit = TRUE)` component are all active simultaneously.  The
resulting ensemble is likely over-dispersed.  Use at most two of the three, or
use `min_window` alone with `boot = NULL` for cleaner diagnostics.

Additionally:

- Innovations are drawn independently per `(unit, time, sim)` row — the engine
  does not model cross-unit common shocks or cross-equation residual
  correlation.
- `parametric_distribution`, `univariate_fable`, and `exogen` carry no
  parameter uncertainty across draws (they are fitted once and shared), unless
  `parametric_distribution` is configured with `param_uncertainty = TRUE`.

See `?endogenr` for the full discussion and roadmap.
